---
title: "R you ready?"
subtitle: "An Introduction to R"
output: html_document
---

## How to follow along and participate

* Open R Studio

* Install the `tidyverse`, `table1`, `haven`, and `here` packages (if you haven't already)

    + Use the `Tools` menu and choose the `Install packages...` menu choice 

    + Type the names of the packages, `tidyverse, table1, haven, here`, in the dialog box that opens and click Install

* Open the workshop R Markdown file and click on the **tiny** green triangle on the top right corner of the gray shaded area (this is a code chunk!) that contains this: 

```{r chunk1}
library(package = "tidyverse")
library(package = "table1")
library(package = "haven")
library(package = "here")
```

* The arrow will run all the code in that "chunk" of code

* Throughout the workshop you will be able to run the code from the demonstration by clicking the arrow on the top right in the corresponding code chunk

* In between code chunks, you can type as you would in a Word document if you'd like to take notes


## Importing data into R


```{r chunk2}
# use here so that R looks in the same folder you are in
here()

# use read_spss function to read the spss file 
teensTechData2018 <- read_spss(file = "March 7-April 10, 2018 - Teens and Tech Survey - SPSS.sav")
```

## Choosing variables

```{r chunk3}
# use select to choose variables
teensTechData2018Rec <- teensTechData2018 %>% 
  select(SNS2, SOC1, OH3_1, OH3_2, OH3_7, AGE, RACETHNICITY, GENDER)
```


## Check your work

```{r chunk4}
summary(object = teensTechData2018Rec)
```

## Choosing variables & people

```{r chunk5}
# use filter to select observations
teensTechData2018Rec <- teensTechData2018 %>% 
  select(SNS2, SOC1, OH3_1, OH3_2, OH3_7, AGE, RACETHNICITY, GENDER) %>% 
  filter(AGE > 14)
```



## Check your work

```{r chunk6}
summary(object = teensTechData2018Rec)
```


## Managing data types in R

* Commonly used data types in R:

    + factor: variable with categories (categorical data) 
    
    + numeric: variable with values over some continuum (continuous data) 
    
    + integer: whole number with values over some continuum 
    
    + character: used for words or other verbatim information (e.g., ZIP codes) 
    
* Several of the variables are the wrong data type! Which ones?
    
* The `mutate()` function is useful for changing variable types or values or really anything 

* We can add this function into our data management chain

## Recoding variables

```{r chunk7}
# recode age and gender
teensTechData2018Rec <- teensTechData2018 %>% 
  select(SNS2, SOC1, OH3_1, OH3_2, OH3_7, AGE, RACETHNICITY, GENDER) %>% 
  filter(AGE > 14) %>% 
  zap_labels() %>% 
  mutate(AGE = as.numeric(AGE)) %>% 
  mutate(GENDER = recode_factor(GENDER,
                                '1' = 'Male',
                                '2' = 'Female',
                                '99' = NA_character_))
```


## Check your work

```{r chunk8}
summary(object = teensTechData2018Rec)
```

## Recoding more variables

```{r chunk9}
# recode race-ethnicity
teensTechData2018Rec <- teensTechData2018 %>% 
  select(SNS2, SOC1, OH3_1, OH3_2, OH3_7, AGE, RACETHNICITY, GENDER) %>% 
  filter(AGE > 14) %>% 
  zap_labels() %>% 
  mutate(AGE = as.numeric(AGE)) %>% 
  mutate(GENDER = recode_factor(GENDER,
                                '1' = 'Male',
                                '2' = 'Female',
                                '99' = NA_character_)) %>% 
  mutate(RACETHNICITY = recode_factor(RACETHNICITY,
                                      '1' = 'White, non-Hispanic',
                                      '2' = 'Black, non-Hispanic',
                                      '3' = 'Other, non-Hispanic',
                                      '4' = 'Hispanic',
                                      '98' = NA_character_)) 
```


## Check your work

```{r chunk10}
summary(object = teensTechData2018Rec)
```


## Recoding tech variables

```{r chunk11}
teensTechData2018Rec <- teensTechData2018 %>% 
  select(SNS2, SOC1, OH3_1, OH3_2, OH3_7, AGE, RACETHNICITY, GENDER) %>% 
  filter(AGE > 14) %>% 
  zap_labels() %>% 
  mutate(AGE = as.numeric(AGE)) %>% 
  mutate(GENDER = recode_factor(GENDER,
                                '1' = 'Male',
                                '2' = 'Female',
                                '99' = NA_character_)) %>% 
  mutate(RACETHNICITY = recode_factor(RACETHNICITY,
                                      '1' = 'White, non-Hispanic',
                                      '2' = 'Black, non-Hispanic',
                                      '3' = 'Other, non-Hispanic',
                                      '4' = 'Hispanic',
                                      '98' = NA_character_)) %>% 
  mutate(SNS2 = recode_factor(SNS2,
                              '1' = 'Twitter',
                              '2' = 'Instagram',
                              '3' = 'Facebook',
                              '4' = 'Snapchat',
                              '5' = 'YouTube',
                              '6' = 'Tumblr',
                              '7' = 'Reddit',
                              '8' = 'Not a social media user',
                              '98' = NA_character_)) %>% 
  mutate(OH3_1 = recode_factor(OH3_1,
                               '0' = 'No',
                               '1' = 'Yes'))
  
```

## Check your work

```{r chunk12}
summary(object = teensTechData2018Rec)
```


## You Try It!

* Finish the recoding yourself by cleaning SOC1, OH3_2, OH3_7

* The codebook is unclear, so here are the labels to add:

    + SOC1: 1 = Mostly positive, 2 = Mostly negative, and 3 = Neither positive nor negative 
    
    + OH3_2 and OH3_7: 1 = Yes, 0 = No
    
* When finished adding this code, run the `summary()` function to see if it works


```{r chunk13}
# add the last 3 variables to the end of this data management code
# remember to use the pipe to move from one recode to the next
teensTechData2018Rec <- teensTechData2018 %>% 
  select(SNS2, SOC1, OH3_1, OH3_2, OH3_7, AGE, RACETHNICITY, GENDER) %>% 
  filter(AGE > 14) %>% 
  zap_labels() %>% 
  mutate(AGE = as.numeric(AGE)) %>% 
  mutate(GENDER = recode_factor(GENDER,
                                '1' = 'Male',
                                '2' = 'Female',
                                '99' = NA_character_)) %>% 
  mutate(RACETHNICITY = recode_factor(RACETHNICITY,
                                      '1' = 'White, non-Hispanic',
                                      '2' = 'Black, non-Hispanic',
                                      '3' = 'Other, non-Hispanic',
                                      '4' = 'Hispanic',
                                      '98' = NA_character_)) %>% 
  mutate(SNS2 = recode_factor(SNS2,
                              '1' = 'Twitter',
                              '2' = 'Instagram',
                              '3' = 'Facebook',
                              '4' = 'Snapchat',
                              '5' = 'YouTube',
                              '6' = 'Tumblr',
                              '7' = 'Reddit',
                              '8' = 'Not a social media user',
                              '98' = NA_character_)) %>% 
  mutate(OH3_1 = recode_factor(OH3_1,
                               '0' = 'No',
                               '1' = 'Yes'))  

  
```

```{r chunk14}
# check your work
summary(object = teensTechData2018Rec)
```


## Basic table

```{r chunk15}
# basic table with 4 variables
table1(~ SOC1 + AGE + GENDER + RACETHNICITY, 
       data = teensTechData2018Rec)
```


## Adding labels & units to the table

```{r chunk16}
label(teensTechData2018Rec$SOC1) <- "What effect would you say social media has had on people your age?"
label(teensTechData2018Rec$RACETHNICITY) <- "Race & ethnicity"
label(teensTechData2018Rec$GENDER) <- "Sex"
label(teensTechData2018Rec$AGE) <- "Age"

units(teensTechData2018Rec$AGE) <- "years"
```


## Fancier table

```{r chunk17}
table1(~ SOC1 + AGE + GENDER + RACETHNICITY, 
       data = teensTechData2018Rec,
       render.missing = NULL,
       render.continuous = c(. = "Mean (SD)"))
```

## Stratified table

```{r chunk18}
table1(~ SOC1 + AGE + RACETHNICITY | GENDER, 
       data = teensTechData2018Rec,
       render.missing = NULL,
       render.continuous = c(. = "Mean (SD)"))
```

## Stratified table

```{r chunk19}
table1(~ SOC1 + AGE | RACETHNICITY * GENDER, 
       data = teensTechData2018Rec,
       overall = NULL,
       render.missing = NULL,
       render.continuous = c(. = "Mean (SD)"))
```



## Basic bar chart

```{r chunk20}
teensTechData2018Rec %>%
  ggplot(aes(x = SNS2)) +
  geom_bar() 
```



## Fancy bar chart

```{r chunk21}
teensTechData2018Rec %>%
  drop_na(SNS2) %>% 
  ggplot(aes(x = SNS2)) +
  geom_bar(fill = "#7463AC") +
  labs(x = "Social media platform",
       y = "Number who use this platform most",
       title = "Most used social media platforms") +
  theme_minimal()
```


## You try it!

Make a bar chart of one of the other variables in your data set with the Figure 1 code in your R Markdown file.

Edit the code to: 

* Change the name of the variable you are plotting 

* Change the color of the bars by replacing #7463AC the name of your favorite color (e.g., orange) or using a color found on the https://colorbrewer2.org/ website

* Edit the labels for the axes and title inside the quote marks

* Try another theme, you can find them here: https://ggplot2.tidyverse.org/reference/ggtheme.html

When you have completed the changes, click the little green arrow on the top right of the code chunk to run the code.

```{r chunk22}
# paste the code from chunk 21 here and revise

```


## Graph with facets

```{r chunk23}
teensTechData2018Rec %>%
  drop_na(OH3_2, GENDER) %>% 
  ggplot(aes(x = OH3_2)) +
  geom_bar(fill = "dodgerblue") +
  labs(x = "Been physically threatened online or on cell phone",
       y = "Frequency",
       title = "Number of participants who have been physically threatened online\nor on cell phone") +
  theme_minimal() +
  facet_wrap(facets = vars(GENDER))
```


## The end

* Email me at harrisj@wustl.edu

* Connect with me on Twitter @jenineharris
